---
title: Factors that affect people's age having their first baby based on cleaned GSS
  data.
author: "Zilong Yi"
date: "19/10/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Factors that affect people's age having their first baby based on cleaned GSS data. 

# Zilong Yi 
# 19/10/2020
# Code and data supporting this analysis is available at: https://github.com/ZilongYi/Ps2

## Abstract
```{r echo=FALSE,message=FALSE}
library(tidyverse)
library(survey)
library(brms)
gss <- read_csv("gss.csv")
data <- gss %>%
  select(age_at_first_birth,age,self_rated_mental_health,sex,
         income_family,feelings_life,province) %>% 
  filter(!is.na(age_at_first_birth)) %>% 
  tibble()
iteration <- c(1:length(data$age_at_first_birth))
data$caseid = iteration
data$Age = data$age - 15
```

As time goes by, tendency of people having first baby later is becoming more popular. The factors contributing to this phenomenon could be gender, mental health,etc. Essentially, this is a complicated topic, however, only a few factors are considered in this report. By using stratified sampling among 10 provinces in Canada, GSS data was obtained and a generalized linear model is performed to analyze the topic. It turns out only a few factors listed,age, gender, income of family,correlated to age of people having first baby. 

## Introduction

The purpose of this report is to give insight on the relation between age of people having their first baby and factors that could affect it. By inspecting given data set, 5 protential factors are selected, i.e. sex, mental health, age, income, feelings about life, to do a generalized linear regression model. After modelling, a few significant varibales have been found and itshould gives enough sense or strong enough model to predict the time of a person with certain characteristics to have his/her first baby.

## Data
```{r, echo=FALSE,message=FALSE}
ggplot(data, aes(x=Age, y = age_at_first_birth )) + geom_point() + labs(title = 'Figure 1')
```

Data used in this report is a subset of 2017 Genearl Social Survey(GSS) on Family. It is most appropriate to choose this data set on family for report as the goal is to analyze factors that affect people's age of having first baby. Also, GSS applied stratified sampling to collect data via list of telephone in use and the Address Register,which ensures variety and interity of data. (more information about how to find data is in the discussion section.)

It is expected that giving birth of one's family involves many complecated factors. However, by inspecting the given data set and ignoring varibales with most entries being NA, it is suspected 5 variables as listed in introduction are correlated to age of haveing first baby. Taking variable age as an example, by Figure 1, it is natural to suspect a linear relationship between age and age_at_first_birth. Similar to other variables to see linear relationships. Also, entries of NA of variable age having first baby have been ignored. 
Clearly, ignoring varibales with most entrise being 0 indicates one of potential drawbacks to the data as some of them may correlate to the goal. Another potential drawback could be potential family being duplicated as GSS used the Address Register and List of Telephone to gather data, and one might have different address associated with Telephone number. Therefore, it could be possiblly counted more than once. 

## Model
```{r, echo=FALSE, message=FALSE}
prov <- unique(data$province)
value <- c(1:10)
coding_table <- tibble(prov, value)
num <- c()
for(i in iteration){
  aim_column <- data %>% filter(caseid == i) 
  n <- aim_column$province
  coding_row <- coding_table %>% filter(prov == n)
  m <- coding_row$value
  num <- append(num, m)
}

data$prov_coding = num
data <- data %>% arrange(prov_coding)
data <- subset(data, select= -caseid)
data$caseid = iteration
size <- c(max(filter(data,prov_coding == 1)$caseid),
          max(filter(data,prov_coding == 2)$caseid)-max(filter(data,prov_coding == 1)$caseid),
          max(filter(data,prov_coding == 3)$caseid)-max(filter(data,prov_coding == 2)$caseid),
          max(filter(data,prov_coding == 4)$caseid)-max(filter(data,prov_coding == 3)$caseid),
          max(filter(data,prov_coding == 5)$caseid)-max(filter(data,prov_coding == 4)$caseid),
          max(filter(data,prov_coding == 6)$caseid)-max(filter(data,prov_coding == 5)$caseid),
          max(filter(data,prov_coding == 7)$caseid)-max(filter(data,prov_coding == 6)$caseid),
          max(filter(data,prov_coding == 8)$caseid)-max(filter(data,prov_coding == 7)$caseid),
          max(filter(data,prov_coding == 9)$caseid)-max(filter(data,prov_coding == 8)$caseid),
          max(filter(data,prov_coding == 10)$caseid)-max(filter(data,prov_coding == 9)$caseid))

N = 30740074
fpc.srs = rep(7084565,size[1])
fpc.srs = append(fpc.srs, c(rep(11965023,size[2]), rep(3493200,size[3]),rep(821282,size[4]),
                            rep(939283,size[5]),rep(648745,size[6]),
                            rep(1087883,size[7]), rep(453724,size[8]),
                            rep(4118671,size[9]),rep(127698,size[10])))
weight = fpc.srs/N
birth.design <- svydesign(id=~1, strata = ~province, weights = ~weight, data=data, fpc = fpc.srs)
mysvylm <- svyglm(age_at_first_birth ~ Age + income_family +self_rated_mental_health 
                  + as.factor(feelings_life) + sex, birth.design)
#summary(mysvylm)
knitr::kable(broom::tidy(mysvylm))
#broom::tidy(mysvylm)
par(mfrow=c(2,2))
plot(mysvylm)
#brm_model <- brm(formula = age_at_first_birth ~ Age + income_family +
#                  self_rated_mental_health + as.factor(feelings_life), data = data, seed = 853 )
#summary(brm_model)
```

The selected model here is generalized linear model(GLM), as dependent variable is continueous. It is assumed there is no interactions among varaibles.
The model could be expessed as $$y_i = \beta_0 + x_{1i}*\beta_1 + x_{2i}*\beta_2 + x_{3i}*\beta_3 + + x_{4i}*\beta_4+ x_{5i} *\beta_5 + \epsilon_i$$
where $y_i$ is the age of peopel giving their first birth, $x_{1i}$ is age of unit i, $x_{2i}$ is the income of $i^{th}$ family, $x_{3i}$ is self-rated mental health of $i^{th}$ unit, $x_{4i}$ is the feeling about life of $i^{th}$ unit and $x_{5i}$ is the gender of $i^{th}$ unit. $\beta_i$s are corresponding coefficients of independent variable $x_i$. $\epsilon_i$ is the error of $i^{th}$ unit and $\beta_0$ is the intercept of linear model representing base case of model. These coeffcients can be calculated in R language with library survey. 

Note, the baseline of model is female people with age 15, 100,00 to 124,999 dollar income of family, could not give rate of their mental health,and 0 out of 10 on feeling about life. Also, independent variables in model contain categorical variable, which means each level should be interpreted indivisually. 

When GLM is used, it is assumed that $\epsilon_i$ is normally distributed with mean 0 and constan varaiance $\sigma^2$. By residual vs fitted plot, mean of residual is around 0, which means the condition of  mean of $\epsilon_i$ equals to 0 is satisfied. 
By normal Q-Q plot, it appears that $\epsilon_i$ does not follow normal distribution, but this is not the case. Since data set used has ignored those entries with value NA of age having first baby, which means GSS did not get response of variable with that particular person. In general, if total population size and sample size has been narrowed down with less NA response, $\epsilon_i$ would normally distributed. 
By scale location and residual leverage plots, conditon that $\epsilon_i$ has constant variance $\sigma^2$ is roughly satisfied, as there is obvious trend in scale location plot and no particular outliers. 

Since it is assumed no interactions between variables, alternative appropriate model would be a generalized linear model with interactions, which would result in a more complicated model but gives more information abou the date. 

## Results
By given coefficient table, each coefficient, i.e. $\beta_i$, has been calculeted or modeled. Thus aiming model would be 
$$\hat{y_i} = 30.713 + (-0.04)*x_{1i}* + x_{2i}*\beta_2 + x_{3i}*\beta_3 + + x_{4i}*\beta_4+ x_{5i} *\beta_5 $$
Note variable $x_{i}\ for\  i=2,3,4,5 $ are categorical variables, it is expected to have different coeffcients for different levels. Taking income_family as an example, for level of 125,000 and more, the coeffcient is 1.02010907.

By indication of p-value, not every variable or every level of a categorical variable is significant. significant variables are age,income of family and gender as each p-value of varibale or of each level of variable is less than 0.01. Similarly, some levels or all level of variable self-rated mental health and feeling about life are not significant with p-value less than 0.05. Different standard of p-value indicates how strong the data support conclusion in terms of significance. 

## Discussion
interpretation of Result:
Since variable $x_{i}\ for\  i=2,3,4,5 $ are categorical variables, it is expected to interprete each level individually. For instance, model predicts that male with age 20, 125,00 and more dollar income of family, Excellent mental health,and 8 out of 10 on feeling about life would having his first baby at age $30.713-0.04*5+1.02-0.05-1.15+2.55 = 32.883$. Other interpretations follows similarly as given example.
Also, note significant variable gender has coeffcient 2.61446512 $\gt 0$, which means males tends to have their first baby later than female by 2.6 years in terms of age. Interestingly, based on the given data set, in general, only family with income 125,000 and more dollar would delay their time to have first baby.
Also, we should expect there is a considerable error in the model given the number of data omitted. 

Discussion of questionnaire:
GSS collected data by phone calls. Clearly, benefits of that are following:
1.response would be recorded immediately,
2.interviewers could explain at real time if there is a confusion about the questions or it is not clear of what is being asked, 
3. Giving more precise definition of questions. For example, questionnaire has question about rating of life, which does not have a precise definition, i.e. definition varies from person to person. In this case, talking to people could have higher chance to get more concrete result/data.

However, this way of gathering information has its own drawbacks:
1. experimental unit might not response to the phone call or has to wait until available, which means time consuming.
2. it should be considered that youngsters are more likely not to answer the phone as it is more popular to use phone for social media. 

Discussion of Methodology:
Also, GSS applied stratified sampling to collect data via list of telephone in use and the Address Register,which ensures variety and interity of data.  
GSS treats each of the 10 provinces as a strata with expection of Census Metropolitan Areas(CMAs) including Toronto, Quebec City, Ottawa, etc. However, after cleaning the data, those CMAs haven been excluded. The target population of GSS is all persons 15 years of age and older in Canada and target sample size is 20,000. A simple random sample without replacement is performed on each stratum. However, stratified sampling assumes homogeneous property of each startum, which may not be the case in real world, i.e. different province may have its unique characteristics. By this assumpting, data losses,to some extent, variety among provinces.
Everytime when there is no repsonse or person who refuses to participate, interviewer would re-contact up to 2 more times explaining importance of the survey and encouraging to participate.

# Weaknesses
1. Calculation of N, population size. 
In this report, N is not exact same as number given by census of Canada, even though it should be. N is a rounded number of each province population in 2017. 
2.Normality of $\epsilon_i$
It is assumed that assumption of GLM is satisfied. However, due to ignoration of experimental unit with NA value in variable age_at_first_birth, normality failed, but instead it is expected to be normal with more response value. 
3. Assumption of homogeneous strata.
GSS collected data using stratified sampling, which assumes homogeneous property between each stratum. However, this may not be true in real world. Different province could have different population composition, which would lead to different characteristics of response in terms of goal of this report.

4. interactions between varaibles.
In this report, it is assumed that variables considered do not interact with each other. However, this is may not be the case. 

# Next Steps
Given weaknesses above, this section would include a few next steps that could be improved in the future study.
1.For population number N, get it from census of Canada. 

2.Increasing target sample or do survey promotion to encourage people to participate in to get more responded value, which could essentially solve issues with nomality of $\epsilon_i$

3.If further prediction required in terms of specific geographic area, specific sampling survey should be conducted.

4. Interactions of variables should be considered in the case where there is no evidence indicating no interactions. 

## References
1.Government of Canada, S. (2017, September 27). Annual Demographic Estimates: Canada, Provinces and Territories Analysis: Population by age and sex. Retrieved October 19, 2020, from https://www150.statcan.gc.ca/n1/pub/91-215-x/2017000/sec2-eng.htm

2. 2017. Public Use Microdata File Documentation And User’S Guide. [ebook] Canada: authority of the Minister responsible for Statistics Canada, pp.8-19. Available at: <https://sda-artsci-utoronto-ca.myaccess.library.utoronto.ca/sdaweb/dli2/gss/gss31/gss31/more_doc/GSS31_User_Guide.pdf> [Accessed 19 October 2020].

3.Generalized linear model. (2020, September 26). Retrieved October 19, 2020, from https://en.wikipedia.org/wiki/Generalized_linear_model

4. Wickham, H., &amp; Grolemund, G. (2017). R for data science: Import, tidy, transform, visualize and model data. Beijing: O'Reilly.

5. Smith, Tom W., Davern, Michael, Freese, Jeremy, and Morgan, Stephen L., General Social Surveys, 1972-2018 [machine-readable data file] /Principal Investigator, Smith, Tom W.; Co-Principal Investigators, Michael Davern, Jeremy Freese and Stephen L. Morgan; Sponsored by National Science Foundation. --NORC ed.-- Chicago: NORC, 2019.
1 data file (64,814 logical records) + 1 codebook (3,758 pp.). -- (National Data Program for the Social Sciences, no. 25).

6. Rohan A., Sam Caetano (2020) gss_cleaning [R].

7. Get Documentation. (n.d.). Retrieved October 20, 2020, from http://gss.norc.org/Get-Documentation

8. Get the Data. (n.d.). Retrieved October 20, 2020, from http://gss.norc.org/Get-The-Data
